---
title: "Homework 7"
author: Akshay Gona
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Problem 1: The infamous mule kick data <small>20pts</small>

The file `mule_kicks.csv`, available for download (here)[https://kdlevin-uwstat.github.io/STAT340-Fall2021/hw/03/mule_kicks.csv], contains a simplified version of a very famous data set. The data consists of the number of soldiers killed by being kicked by mules or horses each year in a number of different companies in the Prussian army near the end of the 19th century.

This may seem at first to be a very silly thing to collect data about, but it is a very interesting thing to look at if you are interested in rare events. Deaths by horse kick were rare events that occurred independently of one another, and thus it is precisely the kind of process that we might expect to obey a Poisson distribution.

Download the data and read it into R by running

```{r}
download.file('https://kdlevin-uwstat.github.io/STAT340-Fall2021/hw/03/mule_kicks.csv', destfile='mule_kicks.csv')
mule_kicks = read.csv('mule_kicks.csv', header=TRUE)

head(mule_kicks)
```

`mule_kicks` contains a single column, called `deaths`.
Each entry is the number of soldiers killed in one corps of the Prussian army in one year.
There are 14 corps in the data set, studied over 20 years, for a total of 280 death counts.


### Part a: estimating the Poisson rate <small>5pts</small>

Assuming that the mule kicks data follows a Poisson distribution, produce a point estimate for the rate parameter $\lambda$.
There are no strictly right or wrong answers, here, though there are certainly better or worse ones.

```{r}
lambdahat <- mean(mule_kicks$deaths) 
lambdahat
```


### Part b: constructing a CI <small>10pts</small>

Using everything you know (Monte Carlo, CLT, etc.), construct a confidence interval for the rate parameter $\lambda$.
Explain in reasonable detail what you are doing and why you are constructing the confidence interval in this way (a few sentences is fine!).

```{r}
sigma_hat <- sqrt(lambdahat)
SE <- sigma_hat / sqrt(length(mule_kicks$deaths))
confLevel <- 0.95
zScore <- qnorm((1 + confLevel) / 2)
MOE <- zScore * SE
confInterval <- c(lambdahat - MOE, lambdahat + MOE)
confInterval
```

***

Standard error multiplied by the z-score for 0.95 to get the margin of error. The confidence interval is then defined as the point estimate +/- the MOE.

***


### Part c: assessing a model <small>5pts</small>

Here's a slightly more open-ended question.
We *assumed* that the data followed a Poisson distribution.
This may or may not be a reasonable assumption.
Use any and all tools that you know to assess (either with code or simply in words) how reasonable or unreasonable this assumption is.

Once again, there are no strictly right or wrong answers here.
Explain and defend your decisions and thought processes in a reasonable way and you will receive full credit.

***

TODO: Code, plots and/or explanation go here.
```{r}
library(ggplot2)

hist <- ggplot(mule_kicks, aes(x = deaths)) +
  geom_histogram(binwidth = 1, fill = "darkgreen", color = "black") +
  labs(title = "Deaths by Mule Kicks",
       x = "Deaths",
       y = "Frequency")

hist
```

Decrease in frequency, can't follow poission.



## Problem 2: Closing the loop <small>10 pts</small>

In our discussion of the Universal Widgets of Madison company from lecture, we said that we were interested in two questions:

1. Estimating the probability $p$ that a widget is functional.
2. How many widgets should be in a batch to ensure that (with high probability) a batch ships with at least $5$ functional widgets in it?

We discussed question (1) at length in lecture.
What about question (2)?
Our client wants to know how many widgets should ship in each batch so as to ensure that the probability there are at least $5$ functional widgets in a batch is at least $0.99$.

Now, suppose that we have observed data and estimated $p$ to be $0.82$.

Use everything you know so far in this course to give a recommendation to the client.
Be sure to explain clearly what you are doing and why.
If there are any steps, assumptions, etc., that you are not 100% pleased with, feel free to point them out.

__Note:__ there are at least two "obvious" ways to solve this problem. One is based on using Monte Carlo (i.e., assume $p=0.82$ is the truth, and try generating batches of different sizes, etc.).
The other uses direct computation of probabilities, using basic facts about Binomial RVs.
Neither of these is necessarily better than the other, and you do not need to use both approaches to receive full credit.
Indeed, you are free to try doing something else entirely, if you wish.
Just explain clearly what you are doing and why!

```{r}
p <- 0.82
desired <- 0.99

simBatch <- function(batch_size, p) {
  widgets <- sum(rbinom(batch_size, 1, p))
  if (widgets >= 5) {
    return(1)
  } else {
    return(0)
  }
}

batches <- seq(1, 100, by = 1)
probabilities <- sapply(batches, function(x) mean(replicate(10000, simBatch(x, p))))

size <- min(batches[probabilities >= desired])
size
```


We have probability as .82, and desired prob as .99. We then define a function  that simulates a batch of widgets, calculates the probability that at least 5 widgets are functional. Next, we perform a Monte Carlo simulation by iterating over different batches of various sizes and calculating the average probability of having at least 5 functional widgets in each batch. NMC = 10000. We get a recommended size from finding the smallest batch size for which there is a probability >= desired probability.


