---
title: "Homework 5"
output: html_document
author: Akshay Gona
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem \#1: Testing coin flips <small>(6 pts)</small>

In the six sequences below, only one of them is actually **randomly generated from independent flips of a fair coin**. Use a combination of everything you know (common sense, Monte Carlo, hypothesis testing, etc.) to identify which is actually random and explain your reasoning.

(For full points, conduct a formal test and report a p-value for each sequence. You may use a combination of multiple tests to arrive at your answer. If you cannot compute a p-value for each sequence, you can still earn a significant amount of partial credit by carefully explaining your reasoning and response as best as you can.)

My advice is **be creative** with the test statistics you come up with to eliminate each sequence! Think of some way of summarizing a sequence of flips that might be useful for comparing against a simulated sequence of random flips. After you come up with an idea for a statistic, remember to run it on many MC generated completely random flips to produce a distribution under the null, which you can then compare with your data to get a p-value. Also, be careful of now you define "more extreme" than the data.

(2 bonus points available if you can find a single test that is powerful enough to reject all the fake sequences together in one step. Yes, at least one such possible test exists.)

```{r}
flips1 = "HTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHTHT"

flips2 = "HHHTHTTTHHTHHTHHHTTTTHTHTHHTTHTHHHTHHTHTTTHTHHHTHTTTHTHTHHTHTHTTHTHHTHTHTTTHTHHHTHTHTTHTHTHHTHTHTHHHTHTTTHTHHTHTHTHHTTTHTHHTHHTTTTHTHTHHHTHTTHTHHTHTHTTHTHHTHTHHHTHHHTHTTTHTTHTTTHTHHHTHTHTTHTHHTHHTHTTT"

flips3 = "HHTHTHTTTHTHHHTHHTTTHTHHTHTTTHTHTHHTHTHTTHTHHHHHHTTTHTHTHHTHTTTHTHHTHTHTTTHTHHHTTHTTTHTHTHHHHTHTTHHTTTTTHTHHHTHTHTTTTTHHHTHHTHHTHHHTTTTHTHTHHHTHHTTTTTHTHHHTHTHTHTTTHTHHHTHTHTHTTHTHHTHTHTHTTTTHTHHHTHTH"

flips4 = "HTHHHHHHHTHTTHHTTHHHTHTHTTTHHTHHHTHHTTHTTTTTTTTTHTHHTTTTTHTHTHTHHTTHTTHTTTTTHHHTHTTTHTHTHHHTHTTTTHTHTHHTTHTHTTHHTHTHHHHTHTTHHTTHTTHTTHTHHHHHHTTTTTTHHHTTHTHHHHTTTHTTHHHTTHTHHTTTHHTHHTTTHTHHTHHHTHHTTHHH"

flips5 = "HHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTTHHHHHHHHHHTTTTTTTTTT"

flips6 = "TTHTTTHTTTTTTTHTHTHTHTTHTTHTHHTHHTTTHHTHTTTHTHHTHHHTHTTHHTHHTTHTHTTTTHTHTTTHHTTTTTTTTHTHHTTHTTTTTTHTHTHTHTTTHTTHHTTHTTTHHTTTHTTHTTTTHTTTTHHTTTHTHTHHHTTTTTTHTHHTTTTTTTTTTTTHHHTTTHHHTTTHTTTHTHTTHTTTTTHT"

# you can use the function below to split the above sequences in vectors of flips
split = function(str) strsplit(str, split="")[[1]]
split(flips1)


count_runs <- function(sequence) {
  runs <- integer()
  current_run <- 1
  for (i in 2:length(sequence)) {
    if (sequence[i] == sequence[i - 1]) {
      current_run <- current_run + 1
    } else {
      runs <- c(runs, current_run)
      current_run <- 1
    }
  }
  runs <- c(runs, current_run)
  return(sum(runs > 1))
}

runs_test <- function(sequence, num_simulations = 10000) {
  observed_runs <- count_runs(sequence)
  random_runs <- numeric(num_simulations)
  for (i in 1:num_simulations) {
    random_runs[i] <- count_runs(sample(c("H", "T"), length(sequence), replace = TRUE))
  }
  p_value <- mean(random_runs >= observed_runs)
  return(p_value)
}

R <- sapply(list(split(flips1), split(flips2), split(flips3), split(flips4), split(flips5), split(flips6)), runs_test)
complement <- 1 - R
complement
```

If we use the test statistic of run length, we find that we can reject all of
the sequences except for the last one, which is significantly above the significance level. Therefore, the last one is the only fair one.

## Problem \#2: Finding the Trick Coin <small>(6 pts; 2pts each)</small>

I have two coins in my pocket - a trick coin with two heads and a fair coin with one head and one tail(s?). We'll play a game. I will grab one coin at random, and flip it $N$ times. After that you will have to decide if it is the fair coin or the trick coin. The null hypothesis is that it is the fair coin. 

**Decision Rule 1**: If after $N$ flips there are no tails, then you decide it is the trick coin. If there is at least 1 tail then you know it is the fair coin. 

a. Using "Decision Rule 1", what is the lowest number of flips $N$ would you need in order to have a significance level less than 5% for this test?
```{r}
N <- (log(0.05) / log(0.5))
ceiling(N)
```

b. Using $N$ from part a, what is the power of the test?
```{r}
power <- 1
power
```

c. Suppose $N=4$ is decided. How can you modify the decision process to have a significance level of exactly 5%? Does this change the power of the test?

New Decision Rule:If all 4 flips are heads, decide it's the trick coin. Otherwise, it's the fair coin. 

Probability of observing all heads in fair coin (P(HHHH)): (1/2)^4 = 1/16. New significance level: 1/16 = 0.0625, which is greater than 0.05. This modified decision rule does not achieve a significance level of exactly 5%. It's slightly more forgiving (α = 0.0625). Power of the test with α = 0.0625: Power = 1 - α = 1 - 0.0625 = 0.9375

The power of the test has slightly decreased to 93.75%. This means there's a slightly lower chance of correctly identifying the trick coin compared to the previous decision rule.

d. Extra Credit (2 points): Suppose if you guess correct you win \$100 (and if you're wrong you get nothing), but each flip of the coin costs \$10. What strategy would you use to maximize your expected profit from this game?

Expected Profit (E): The average profit you expect to make per game.
Cost per flip: -$10
Reward for correct guess: $100

Flip the coin once: This costs $10.
If it's heads, guess the trick coin. There's a 50% chance of being correct (fair coin or trick coin).
Correct guess: Gain \$100 - \$10 (flip cost) = $90 profit.
Incorrect guess: Gain $0 (no reward).
If it's tails, guess the fair coin. You're guaranteed to be correct. Gain \$100 - \$10 (flip cost) = $90 profit.



## Problem \#3: Testing the maximum of a uniform distribution <small>(8 pts; 2 pts each)</small>

We sample $X_1, X_x,\ldots,X_n \overset{\text{iid}}\sim\text{Uniform}(0,m)$ where $m$ is an unknown maximum. Sleazy Jim tells you that $m=1$ but you're not so sure. The 50 values sampled are in the following data file:

```{r}
X <- read.csv("uniform_sample.csv")$x
```

a. Write out in formal notation the null and alternative hypotheses. 

$H_0 : m = 1$ maximum value (m) of the uniform distribution is equal to 1.

$H_0 : m \not=1$ maximum value (m) of the uniform distribution is not equal to 1.

b. Come up with a test statistic and measure your sampled data. Is this a one-sided test or two-sided test?

We can check the maximum value in the data set. If its greater than 1, then there is evidence against the null.
```{r}
Xmax = max(X)
Xmax
```
Since we're interested in whether m is different from 1 in either direction, it's a two-sided test.

c. Simulate a distribution for the test statistic under the null hypothesis of size at least 1000. Display a histogram of your test statistic distribution.

```{r}
n_simulations <- 1000
max_values <- replicate(n_simulations, max(runif(length(X), 0, 1)))
hist(max_values, breaks = 20, main = "Distribution of Maximum Values under H0", xlab = "Maximum Value")
```
d. Calculate the $p$-value for this data and make a conclusion.

```{r}
p_value <- 1 - mean(max_values >= Xmax) + mean(max_values <= 1 - Xmax)
p_value
```

Fail to reject the null hypothesis. There is not enough evidence to conclude that the maximum value is different from 1. P value is greater than 0.05.



## Problem \#4: Rising Temperatures? <small>(10 pts; 2 pt each)</small>

The `annual_avg_temp.csv` data file contains the US annual average temperature from 1875 through 2022.
```{r}
temp <- read.csv("annual_avg_temp.csv")
temps <- temp$Annual.Average.Temperature.F
plot(temp, type="l")
```

There seems to be a trend but it could be due to randomness. Your task is to perform a permutation test on the historical record of annual avg. temperatures to determine if there is statistical evidence of a real trend.

a. State the null and alternative hypotheses

null hypothesis: trend that seems to appear is due only to randomness. 
alternative hypothesis: there is an increasing trend in the average annual temperature as the year increases.

b. Determine a test statistic that identify non-randomness in the temperatures

average temperature increase between consecutive years

c. Decide whether the test will be a one or two-tailed test

one-tailed test.

d. Simulate a distribution of test statistics under the null hypothesis
```{r}
differences <- diff(temp$Annual.Average.Temperature.F)
observed_statistic <- mean(differences)
observed_statistic

NMC <- 10000
permuted_statistics <- numeric(NMC)

for (i in 1:NMC) {
  simul <- sample(temp$Annual.Average.Temperature.F)
  differences <- diff(simul)
  permuted_statistics[i] <- mean(differences)
}
```

e. Calculate the test statistic on the observed data, calculate the $p$-value and state your conclusions.

```{r}
p_value <- mean(permuted_statistics >= observed_statistic)
p_value
```


*Hint: basing the test statistic on the differences between consecutive years may be a good idea.*

